Virtual Memory Assisted Transactional Memory for Dynamic Languages

### 2014-06-02 talk/dls2014/paper/paper.tex
Tweaks

TM systems come in can be broadly categorised as hardware based (HTM),
software based (STM), or hybrid systems (HyTM). HTM systems are limited
by hardware constraints~\cite{odaira14,fuad10}, while STM systems have
a lot of overhead~\cite{cascaval08,drago11}. In \cite{wayforward14},
we argue that STM is still the best way forward, especially since it
supports large atomic blocks as a new way for synchronising multiple
threads. There have been several attempts at lowering the overhead
of STM~\cite{warmhoff13,spear09} - sometimes at the cost of scalability.
In this paper, we describe how we manage to lower the overhead of our
STM system so that it can be seen as a viable replacement for the GIL.

Our contributions include:
\begin{itemize}[noitemsep]
\item We introduce a new software transactional memory (STM) system
  that performs well even on low numbers of CPUs. It uses a novel
  combination of long-existing CPU features and garbage collector (GC)
  integration in order to keep the overhead of STM very low.
\item This new STM system is used to replace the GIL in Python and is
  then evaluated extensively.
\item We introduce atomic blocks to the Python language to provide a
  backwards compatible, composable synchronisation mechanism for
  threads.
\end{itemize}

@@

Looking back at address translation for object references, we see now
that this is actually a two-step process. First, $\%gs{::}SO$ gets
translated to different linear addresses in different threads by the
CPU. Then, depending on the current mapping of virtual pages to file
pages, these LAs can map to a single file page in the sharing-segment,
or to privatised file pages in the corresponding segments. This
mapping is also performed efficiently by CPUs that have a Memory
Management Unit (MMU) and can easily be done on every access to an
object.

@@

We now use these mechanisms to provide isolation for transactions.
Using write barriers, we implement a \emph{Copy-On-Write (COW)} on the
level of pages. Starting from the initial fully-shared configuration
(figure \ref{fig:Page-Remapping}, (II)), when we need to modify an
object without other threads seeing the changes immediately, we ensure
that all pages belonging to the object are private to our segment.

To detect when to privatise pages, we use write barriers before every
write to an object. When the barrier detects that the object is not in
a private page (or any pages that belong to the object), we remap and
copy the pages to the thread's segment. From now on, the translation
of $\%gs{::}SO$ in this particular thread will resolve to a private
version of the object automatically. Note that the $SO$ used to reference
the object does not change during that process.

>>>>

TM systems come in\arigo{typo?} can be broadly categorised as hardware based (HTM),
software based (STM), or hybrid systems (HyTM). HTM systems are limited
by hardware constraints~\cite{odaira14,fuad10}, while STM systems have
a lot of overhead~\cite{cascaval08,drago11}. In \cite{wayforward14},
we argue that STM is still the best way forward, especially since it
supports large atomic blocks as a new way for synchronising multiple
threads. There have been several attempts at lowering the overhead
of STM~\cite{warmhoff13,spear09} - sometimes at the cost of scalability.
In this paper, we describe how we manage to lower the overhead of our
STM system so that it can be seen as a viable replacement for the GIL.

Our contributions include:
\begin{itemize}[noitemsep]
\item We introduce a new software transactional memory (STM) system
  that performs well even on low numbers of CPUs. It uses a novel
  combination of long-existing CPU features and garbage collector (GC)
  integration in order to keep the overhead of STM very low.
\item This new STM system is used to replace the GIL in one implementation
  of Python and is then evaluated extensively.
\item We introduce atomic blocks to the Python language to provide a
  backwards compatible, composable synchronisation mechanism for
  threads.
\end{itemize}

@@

Looking back at address translation for object references, we see now
that this is actually a two-step process. First, $\%gs{::}SO$ gets
translated to different linear addresses in different threads by the
CPU. Then, depending on the current mapping of virtual pages to file
pages, these LAs can map to a single file page in the sharing-segment,
or to privatised file pages in the corresponding segments. This
mapping is cost-free: it is part of the normal operations done by the
Memory Management Unit (MMU) of the CPU on every memory access.

@@

We now use these mechanisms to provide isolation for transactions.
Using write barriers, we implement a \emph{Copy-On-Write (COW)} on the
level of pages~\footnote{Conflict detection still occurs on the level
of objects.}. Starting from the initial fully-shared configuration
(figure \ref{fig:Page-Remapping}, (II)), when we need to modify an
object without other threads seeing the changes immediately, we ensure
that all pages belonging to the object are private to our segment.

More precisely, this is done by a write barrier that detects that we are
about to write to an old (pre-transaction) object that we did not record
in the write-set yet.  When this occurs, the slow-path of the write barrier
will also check if the page (or pages) containing the object is still
shared, and if so, privatise it.  This is done by remapping and copying
the page to the thread's own segment. From now on, the translation
of $\%gs{::}SO$ in this particular thread will resolve to a private
version of the object automatically. Note that the $SO$ used to reference
the object does not change during that process.
